Traditional agroforestry systems in Timor-Leste can store large amounts of carbon in both soil and biomass

### 2025-04-14 manuscript.Rmd
manuscript after R1

---
title: "Traditional agroforestry systems in Timor-Leste can store large amounts of carbon in both soil and biomass"
author: "Camille Piponiot, Marguerite Cogné, Vincent Freycon, Alexis Thoumazeau, Marçal Gusmão, Régis Peltier"
abstract:   |
  Agroforestry has the potential to make agriculture more resilient while improving carbon sequestration by incorporating trees and/or other woody perennials into agricultural land thereby  diversifying landscapes. Traditional agricultural systems in tropical areas often include trees, but their carbon sequestration potential is not always well described, hindering their inclusion in climate change mitigation strategies. In this study, we quantified carbon storage in both vegetation biomass and soil in five traditional agroforestry systems (AFS) in Timor-Leste, namely cropping systems with fallow (CF), silvopastures (SP), young agroforests (YA), home gardens (HG), and forest gardens (FG). Our results show that these traditional AFS can store large amounts of carbon, with the average being 155\ Mg\ C\ ha$^{-1}$. The AFS with the highest carbon stocks (FG) stored an average of 210\ Mg\ C\ ha$^{-1}$,  close to the values of old-growth tropical forests. Other AFS stored on average: 107\ Mg\ C\ ha$^{-1}$ in CF, 158\ Mg\ C\ ha$^{-1}$ in SP, 132\ Mg\ C\ ha$^{-1}$ in YA, and 167\ Mg\ C\ ha$^{-1}$ in HG.  Biomass carbon stocks varied significantly between AFS (which differed in tree cover), while soil carbon stocks were less variable between AFS but more dependent on site. We found no relationship between the amount of carbon stored in biomass and soil. Our results highlight the high diversity of traditional AFS in Timor-Leste and their high carbon sequestration capacity. These results could provide an important baseline for the inclusion of AFS in Timor-Leste’s climate change mitigation strategy, and could serve as a reference for future AFS studies in different agro-climates of Timor-Leste.

journal: "Agroforestry systems"
date: ""
output:
  officedown::rdocx_document: 
    base_format: "bookdown::word_document2"
    reference_docx: misc/word-styles-ref-02.docx
bibliography: misc/references.bib
nocite: |
csl: misc/springer-basic-author-date.csl
keywords: "Traditional agroforestry, carbon stocks, soil, biomass"
editor_options: 
  markdown: 
    wrap: 72
---

@@

Agricultural expansion has been the main driver of tropical deforestation over the past half century [@Pendrill2022; @Gibbs2010; @Rudel2009], resulting in large greenhouse gas emissions [approximately 2.6 gigatonnes CO~2~e/yr over 2010-2014; @Pendrill2019]. The demand for tropical agricultural land is expected to continue to grow as the world’s population increases [@UN2022]. In addition, tropical areas are expected to be disproportionately affected by climate change, putting their agriculture and food production systems at risk [@Lawrence2015]. Therefore, there is an urgent need to promote tropical agricultural landscapes that are more resilient to climate change while maintaining high productivity and ecological value [@Harvey2014]. Increasing the amount of carbon stored in the soil and vegetation biomass of agricultural landscapes could be an effective way to mitigate climate change while increasing the resilience of food production systems [@Lal2004; @Minasny2017].

Agroforestry has attracted attention in recent decades as a set of effective agricultural practices that could contribute to climate change mitigation and adaptation [@Albrecht2003; @Verchot2007; @Lasco2014; @Cardinael2021]. Agroforestry is a collective name for land use systems and technologies where woody perennials (trees, shrubs, bamboos, creepers, etc.) are deliberately used on the same land management units as agricultural crops and/or animals, in some form of spatial arrangement or temporal sequence [@Lundgren1982]. Therefore, agroforestry practices allow for the diversification of agricultural production, making agricultural systems more resilient to climate change [@Duffy2021; @Terasaki2023].

Agroforestry systems (AFS) have the potential to store large amounts of carbon in the vegetation biomass and soil, and this carbon storage capacity depends on the type of the AFS in addition to environmental conditions [@Albrecht2003; @Feliciano2018]. Since trees can store large amounts of carbon, the carbon stored in vegetation aboveground biomass of AFS is largely dependent on the number and size of trees in the system [@Ma2020]. However, the effect of AFS on soil organic carbon (SOC) is more variable. Although AFS appear to increase SOC stocks overall in tropical regions, some AFS such as woodlots (i.e, planting trees with intercropping during the establishment phase) may decrease SOC stocks compared to non-AFS [@Feliciano2018], while AFS such as alley cropping and home gardens have been shown to increase SOC stocks [@Oelbermann2004;@Shi2018], and other AFS such as silvopastures have mixed results [@Feliciano2018;@Shi2018]. Other factors such as climate, soil characteristics and previous land use type can also influence SOC in AFS, which may explain the variability of observations reported in the literature [@Nair2009;@Feliciano2018;@Ma2020;@Martin2020]. Furthermore, the relationship between the amount of carbon stored in aboveground vegetation and soil in AFS was only evidenced to a depth of 10 cm [@Ma2020].

In Timor-Leste, a large proportion of the population depends on agriculture for their livelihoods. However, population growth, increasing pressure on natural resources, and an increasingly variable climate with an unreliable rainfall regime are threatening food security [@WB2009; @Molyneux2012]. In this context, productive, resilient, and sustainable agriculture is critical for the country and the livelihood of its people. To address this, AFS have been presented as a potential solution to many of the challenges facing Timorese agriculture [@Paudel2022; @Cogne2024]. Although trees already form a large part of traditional agricultural systems in Timor-Leste [@Paudel2022; @Cogne2024] and can be key elements for biodiversity conservation [@Perfecto2008; @Torquebiau1992], they have received little attention and, to our knowledge, no study has quantified carbon storage in traditional Timorese AFS. A precise quantification of these carbon stocks will improve understanding of the ecological importance of AFS at both local and global scales, and help integrate them into national and international climate change mitigation initiatives.

In this study, we aim to better understand traditional AFS in Timor-Leste by quantifying carbon storage in both vegetation biomass and soil of five traditional AFS: cropping systems with fallow, silvopastures, young agroforests, home gardens, and forest gardens. To do this, we quantified carbon stored in biomass and soil in 30 agroforestry plots across 2 sites in the Baucau district of Timor-Leste. We developed a modeling framework to quantify the effect of agricultural practices (here, different types of traditional AFS) on vegetation aboveground carbon (AGC) and SOC stocks, while controlling for climate and soil variability in the study area. Additionally, we quantified the effect of AGC on SOC stocks: in line with @Ma2020, we expected a positive relationship of AGC with SOC in the topsoil layer (0-10 cm), but no relationship with SOC below 10 cm.

@@

site_summary = site_env[, .(
  coord = paste(round(mean(lat),2), round(mean(long),2), sep = ", "), 
  prec = round(mean(prec)), 
  elev = paste(round(quantile(elev, c(0.025, 0.975))/10)*10, collapse = " and ")),
  .(site)]
```

The research was carried out in the center of the Baucau municipality, located on the north-eastern coast of Timor-Leste. Two communities, or *suco* (a group of villages), were selected to represent two contrasting topo-geographical, climatic and historical cases. The first, Gariuai *suco* (Lat-Lon coordinates: `r site_summary[site=="Gariuai", coord]`), is located on a plateau of Baucau limestone, with an annual rainfall of `r site_summary[site=="Gariuai", prec]` mm and an altitude ranging between `r site_summary[site=="Gariuai", elev]` m above sea level (masl). The second, the village of Osso-Luga in Samalari *suco* (Lat-Lon coordinates: `r site_summary[site=="Osso-Luga", coord]`), is located on the Matebian foothills based on Bobonaro Scaly Clay, with an annual rainfall of `r site_summary[site=="Osso-Luga", prec]` mm and an altitude ranging between `r site_summary[site=="Osso-Luga", elev]` masl (Figure \@ref(fig:map-sites)). The soils of the Baucau municipality are classified using the WRB soil classification [@Mantel2023] as Lithosols and Calcic Luvisols [@FAO2012]. The average soil properties differ between the two sites (Table S\@ref(tab:site-soil)); Gariuai has a clay loam texture whereas Osso-Luga has a clay texture. The two villages also have different histories. Gariuai has been inhabited for at least 50 years, while Osso-Luga was abandoned during the Indonesian occupation and was only repopulated after Timor-Leste gained independence in 2002. The AFS in Osso-Luga lay fallow during the period of abandonment. The land use of these 2 villages is typical of the Baucau district. Land use in these two villages is typical of the Baucau district. The 5 AFS described in this article [see detailed description in @Cogne2024] are found in the valleys and high plateaux, while non-AFS such as rainfed rice and horticultural plots are more likely to be found in the lowlands. Other non-agricultural land uses are also present, such as limestone and limestone quarrying. Some natural forest parks are also preserved around mountain peaks, such as Matebian or Mundo Perdido. Villages are concentrated near paved roads and on hilltops, more rarely near rivers because of the risk of flooding.

## Typology of agroforestry systems

We used the typology of AFS described in @Cogne2024, which was developed in the same study area. This typology was based on the one proposed by Nair *et al.* (1993, 2021) and adapted to the local context with semi-structured interviews previously conducted in another study at the same sites [@Cogne2024]. The five AFS identified were as follows: Crop systems including a wooded Fallow phase (CF), Silvopastures (SP), Young Agroforests (YA), Home Gardens (HG), and Forest Gardens (FG). The AFS are presented in more detail in the Supplemental Information (Table S\@ref(tab:afs-typology)) and in @Cogne2024.

Briefly, the CF are plantations of a few non-perennial crops (e.g., maize [*Zea mays*], sweet potatoes [*Ipomoea batatas*], groundnuts [*Arachis hypogaea*], and squash gourd [*Cucurbita moschata*]) which are left fallow after one to three years of cropping. SP are vast areas (generally 200 to 500 ha) where cattle (*Bos taurus*, *Bos domesticus*, *Bubalus bubalis*) are grazed under the shade of palms, like coconut (*Cocos nucifera*), and trees, like tamarind (*Tamarindus indica*) or eucalyptus (*Eucalyptus alba*) and jujube (*Ziziphus sp.*). YA are agricultural plots where farmers plant rows of trees, like teak (*Tectona grandis*) and mahogany (*Swietenia mahagoni*), spaced approximately 5 x 10 m apart, which are generally left to grow to become HG or FG (see Figure S\@ref(fig:seq-diagram)). HG are plots of land located close to houses where crops and livestock (poultry [*Gallus gallus domesticus*], pigs [*Sus domesticus*], cattle, etc.) are kept in the shade of numerous trees, like Breadfruit (*Artocarpus altilis*), candlenut (*Aleurites moluccanus*) and palms, like areca palm (*Areca catechu*), for multiple uses (wood, fruit, fiber, etc.). FG are generally former HG (over 50 years old) which have been abandoned, for example as a result of the war, and are now used mainly for gathering fruits (e.g. mango [*Mangifera indica*]) and non-timber forest products (e.g. candlenut) with limited human influence.

@@

We randomly selected three plots from each of the five traditional AFS identified in each of the two sites (30 plots in total; Figure \@ref(fig:map-sites)). Sampling plots were defined as the entire agricultural plot (area between 0.06 and 1.25 ha) for all AFS except SP plots, which were generally much larger than 1 ha. In SP plots, a 1 ha sampling plot was delineated within the agricultural plot. Tree and soil measurements (described below) were carried out between June and August 2021.

Within each sampling plot, all trees with a circumference at breast height (CBH) greater than or equal to 100 cm were identified taxonomically and their CBH and height were measured. Five subplots of 10x10 m (0.01 ha) were established at each corner and in the center of the sampling plot (Figure S\@ref(fig:design)). Within these subplots, all trees with a CBH between 30 and 100 cm were taxonomically identified and their CBH and height were measured. Height measurements were taken with a dendrometer.

Taxonomic identification was carried out using Tetum (a national official language) vernacular names, which were then linked to scientific names using the iNaturalist and Pl\@ntNet applications, Geoffrey Hull’s monograph [@Hull2006], the “Useful tropical plants” database [@Fern2014] as well as the Australian Northern Territory Government's Native Plants Herbarium [@NT2013]. The correspondence between Tetum names and scientific names is provided in the Supplemental Information (Table S\@ref(tab:tetum-names)).

In each of the 30 sample plots, a composite of three samples was taken from three of the five 10 x 10 m subplots (including the plot center and two opposite corners; Figure S\@ref(fig:design)), covering the intra-plot variability and at a minimum distance of 1 m from any tree. Soils were sampled with a soil auger (Model Bor Tanah - Dormer Standard Soil Auger, Adiguna Karya Persada, Bogor, Indonesia) at 0-10 cm and 20-30 cm depth. As the length of the head auger (15 cm) was longer than our target soil depths (10 cm), we could not accurately sample the 10-20 cm soil layer. Soils samples were then air-dried and sieved at 2 mm. They underwent standard physical and chemical analyses at Cirad laboratories in Montpellier, France (Table S\@ref(tab:site-soil)). Total carbon content was determined by dry combustion, following the ISO 10694:1995 standard method, and measured with a thermal conductive detector (EA 1112 Elemental Analyser, Thermo Fischer Scientific, Netherlands). Carbonate content was determined with a calcimeter, following the ISO 10693 standard method. Soil organic carbon content was calculated as the difference of the total carbon and the carbonate contents. Soil texture (proportion of clay, silt and sand), and pH were measured as xxx.

In addition, at one location in the center of the plot, three 250 cm$^3$ soil samples were taken with a soil sampling cylinder (5 cm in height, 7.98 cm inner diameter, and 8.4 cm outer diameter; Eijkelkamp Soil & Water, Giesbeek, Netherlands)  at depths of 2-7 cm, 12-17 cm and 22-27 cm. These depths were chosen to take account of the height of the cylinder (i.e. 5 cm), the destruction of the soil below the cylinder during its extraction, and to be associated with depths 0-10 cm, 10-20 cm and 20-30 cm, the depths at which the organic carbon content was calculated or estimated. Each sample was dried at 105°C in a classic oven until a stable dry mass was reached and then sieved to 2 mm. The fine fraction (\< 2 mm) was weighed to be used in the SOC stocks calculation (equation \@ref(eq:soc)).

@@

## height in cm, with a conversion factor between 0.8 and 1
data_sim_agc[GENUS %in% palm_genera, 
             HEIGHT := ifelse(
               iter == "mean", 
               HEIGHT*.9* 100,  ## height in cm
               rnorm(length(HEIGHT), HEIGHT, HEIGHT*cvH)*
                 runif(length(HEIGHT), .8, 1)* 100
             )]

## add WD data
data_wd_palms = unique(data_wd[data_wd$genus %in% palm_genera, c("genus", "meanWD", "sdWD")])
data_sim_agc = merge(
  data_sim_agc,
  data_wd_palms,
  all.x = TRUE,
  by.x = "GENUS",
  by.y = "genus"
)
data_sim_agc[GENUS %in% palm_genera, 
             WD := ifelse(iter == "mean", 
                          meanWD, 
                          rnorm(length(meanWD), meanWD,sdWD))]

data_sim_agc[GENUS %in% palm_genera, 
             AGC := WD * HEIGHT * (DIAM_130_CM/2)^2 * pi * 
               0.89 * # volumetric shrinkage
               1e-6 * # convert from g to tons
               0.4713 # conversion factor from biomass to carbon
]
```

We estimated tree-level AGC stocks from the tree CBH and height measurements using the pantropical allometric equation from @Chave2014 and the R package BIOMASS [@Rejou-Mechain2017]. Each tree was assigned a mean and standard deviation of wood density at the species or genus level when at least one value was available from the Global Wood Density Database [@Zanne2009]. Other taxa were assigned a plot-level average wood density. The biomass of the palms was estimated by multiplying their estimated volume by their estimated wood density. The volume of the palm trunks was estimated as a cylinder, using the height and diameter measured *in situ*, with a volumetric shrinkage of 11% [@WD2007]. The height of the palms was measured above the leaves and was therefore multiplied by a conversion factor of 0.9 to obtain the trunk height. Biomass was then converted into carbon with a factor of 0.4713 [@Thomas2012].

Belowground carbon (BGC) stocks were estimated at tree level using the root-to-shoot ratio equation from @Ledo2018, which is based on tree DBH and climatic water deficit from @Chave2014. This ratio was then multiplied by the estimated AGC stock to obtain tree-by-tree BGC stock.

```{r bgc-estimation}
# root-to-shoot ratio allometry from Ledo et al., 2018
# use CWD values from Chave et al., 2014

@@

Plot-level AGC and BGC stocks were then obtained by summing tree-level stocks and dividing by the plot area.

@@

\begin{equation}
SOC_p = \frac{\sum_i Corg_i \cdot (mfine_i \cdot 10^{-6}) \cdot (Splot \cdot \Delta d)}{Vcyl} (\#eq:soc)
\end{equation}

Where $Corg_i$ is the SOC content at depth i (either 0-10, 10-20 or 20-30 cm) as measured in the laboratory from the soil sampled by the auger and estimated at depth 10-20 cm (see details below); $mfine_i$ is the mass of the fine fraction of the soil in the sampling cylinder at depth i (in g), converted into Mg with a factor of 10$^-6$; $Surf\ =\ 10^8$ is the surface conversion factor from cm$^2$ to ha; $\Delta d$ is the difference in sample depths (= 10 cm) and $Vcyl$ is the volume of the sampling cylinder (250 cm$^3$). This calculation corresponds to the method M4 recommended by @Poeplau2017, which correctly considers the coarse fraction of the soil. We considered that $mfine_i$ at depths of 2-7 cm, 12-17 cm and 22-27 cm was associated with $Corg_i$ at depths of 0-10, 10-20 cm and 20-30 cm, respectively. At depth $i$ = 10-20 cm, where the carbon content was not measured (see “Field inventories” section), we estimated SOC content as a function of SOC content at 0-10 cm and 20-30 cm, assuming a standard negative exponential model of SOC content decrease with soil depth [@Mishra2009;@Murphy2019]:

>>>>

---
title: "Traditional agroforestry systems in Timor-Leste can store large amounts of carbon in both soil and biomass"
author: "Camille Piponiot, Marguerite Cogné, Vincent Freycon, Alexis Thoumazeau, Marçal Gusmão, Régis Peltier"
abstract: |
  Agroforestry has the potential to make agriculture more resilient while improving carbon sequestration by incorporating trees and/or other woody perennials into agricultural land, thereby diversifying landscapes. Traditional agricultural systems in tropical areas often include trees, but their carbon sequestration potential is not always well described, hindering their inclusion in climate change mitigation strategies. In this study, we quantified carbon storage in both vegetation biomass and soil in five traditional agroforestry systems (AFS) in Timor-Leste, namely cropping systems with fallow (CF), silvopastures (SP), young agroforests (YA), home gardens (HG), and forest gardens (FG). Our results show that these traditional AFS can store large amounts of carbon, with the average being 155\ Mg\ C\ ha$^{-1}$. The AFS with the highest total carbon stocks (FG) stored an average of 213\ Mg\ C\ ha$^{-1}$. The average stocks stored by the other AFS were 108\ Mg\ C\ ha$^{-1}$ in CF, 158\ Mg\ C\ ha$^{-1}$ in SP, 134\ Mg\ C\ ha$^{-1}$ in YA, and 171\ Mg\ C\ ha$^{-1}$ in HG. Biomass carbon stocks varied substantially between AFS (which differed in tree cover), while soil carbon stocks were less variable between AFS but more site-dependent. We found no relationship between the amount of carbon stored in biomass and soil. Our results highlight the high diversity of traditional AFS in Timor-Leste and their high carbon sequestration capacity. These results could provide an important baseline for the inclusion of AFS in Timor-Leste’s climate change mitigation strategy, and could serve as a reference for future AFS studies in different agro-climates of Timor-Leste.

journal: "Agroforestry systems"
date: ""
output:
  officedown::rdocx_document: 
    base_format: "bookdown::word_document2"
    reference_docx: misc/word-styles-ref-02.docx
bibliography: misc/references.bib
nocite: |
  @Nair1993, @Nair2021
csl: misc/springer-basic-author-date.csl
keywords: "Traditional agroforestry, carbon stocks, soil, biomass"
editor_options: 
  markdown: 
    wrap: 72
---

@@

Agricultural expansion has been the main driver of tropical deforestation over the past half century [@Pendrill2022; @Gibbs2010; @Rudel2009], resulting in large greenhouse gas emissions [approximately 2.6 gigatonnes CO~2~e yr$^{-1}$ over 2010-2014; @Pendrill2019]. The demand for tropical agricultural land is expected to continue to grow as the world’s population increases [@UN2022]. In addition, tropical areas are expected to be disproportionately affected by climate change, putting their agriculture and food production systems at risk [@Lawrence2015]. Therefore, there is an urgent need to promote tropical agricultural landscapes that are more resilient to climate change while maintaining high productivity and ecological value [@Harvey2014]. Increasing the amount of carbon stored in the soil and vegetation biomass of agricultural landscapes could be an effective way to mitigate climate change while increasing the resilience of food production systems [@Lal2004; @Minasny2017].

Agroforestry has attracted attention in recent decades as a set of effective agricultural practices that could contribute to climate change mitigation and adaptation[@Albrecht2003; @Verchot2007; @Lasco2014; @Cardinael2021]. Agroforestry is the collective name for land-use systems and technologies where woody perennials (trees, shrubs, bamboos, creepers, etc.) are deliberately used on the same land management units as agricultural crops and/or animals in some form of spatial arrangement or temporal sequence [@Lundgren1982]. Agroforestry practices thus allow for the diversification of agricultural production, making agricultural systems more resilient to climate change [@Duffy2021; @Terasaki2023]. Agroforestry practices have long been used in traditional agricultural systems for both environmental and economic benefits [@Viswanath2017].

Agroforestry systems (AFS) have the potential to store large amounts of carbon in vegetation biomass and soil. This carbon storage capacity depends on the type of the AFS as well as environmental conditions [@Albrecht2003; @Feliciano2018]. Since trees can store large amounts of carbon, the carbon stored in the vegetation aboveground biomass of AFS is largely dependent on the number and size of trees in the system [@Ma2020]. However, the effect of AFS on soil organic carbon (SOC) is more variable. Although AFS appear to increase SOC stocks overall in tropical regions, some AFS such as woodlots (i.e. planting trees with intercropping during the establishment phase) may decrease SOC stocks, possibly due to the initial soil disturbance caused by tree planting [@Feliciano2018]. Conversely, AFS such as alley cropping and home gardens have been shown to increase SOC stocks [@Oelbermann2004; @Shi2018], while other AFS such as silvopastures have shown mixed results [@Feliciano2018; @Shi2018]. Other factors such as climate, soil characteristics and previous land-use type can also influence SOC in AFS, which may explain the variability of the observations reported in the literature [@Nair2009; @Feliciano2018; @Ma2020; @Martin2020]. Furthermore, the relationship between the amount of carbon stored in aboveground vegetation and soil in AFS has only been evidenced to a depth of 10 cm [@Ma2020].

In Timor-Leste, a large proportion of the population depends on agriculture for their livelihoods. However, population growth, increasing pressure on natural resources, and an increasingly variable climate with an unreliable rainfall regime are threatening food security [@WB2009; @Molyneux2012]. In this context, productive, resilient, and sustainable agriculture is critical for the country and the livelihoods of its people. AFS have been presented as a potential solution to many of the challenges facing Timorese agriculture [@Paudel2022; @Cogne2024]. Although trees already form a large part of traditional agricultural systems in Timor-Leste [@Paudel2022; @Cogne2024] and can be key elements for biodiversity conservation [@Perfecto2008; @Torquebiau1992], they have received little attention and, to our knowledge, no study has quantified carbon storage in traditional Timorese AFS. A precise quantification of these carbon stocks will improve understanding of the ecological importance of AFS at both local and global scales, and help integrate them into national and international climate change mitigation initiatives.

In this study, we aim to better understand traditional AFS in Timor-Leste by quantifying carbon storage in both the vegetation biomass and soil of five traditional AFS: cropping systems with fallow, silvopastures, young agroforests, home gardens, and forest gardens. To do this, we quantified carbon stored in biomass and soil in 30 agroforestry plots on two sites in the Baucau municipality of Timor-Leste. We developed a modelling framework to quantify the effect of agricultural practices (here, different types of traditional AFS) on vegetation aboveground carbon (AGC) and SOC stocks, while controlling for climate and soil variability in the study area. In addition, we quantified the effect of AGC on SOC stocks. In line with @Ma2020, we expected a positive relationship of AGC with SOC in the topsoil layer (0-10 cm), but no relationship with SOC at 0-30 cm depth.

@@

site_summary = site_env[, .(
  coord = paste0(abs(round(mean(lat),2)), "°S, ", round(mean(long),2), "°E"), 
  prec = round(mean(prec)), 
  elev = paste(round(quantile(elev, c(0.025, 0.975))/10)*10, collapse = " and ")),
  .(site)]
```

The research was carried out in the centre of the Baucau municipality, located on the north-eastern coast of Timor-Leste. Two communities, or *suco* (a group of villages), were selected to represent two contrasting topo-geographical, climatic and historical cases. The first, Gariuai *suco* (`r site_summary[site=="Gariuai", coord]`), is located on a plateau of Baucau limestone, with an annual rainfall of `r site_summary[site=="Gariuai", prec]` mm and an altitude ranging between `r site_summary[site=="Gariuai", elev]` m above sea level (masl). The second, the village of Osso-Luga in Samalari *suco* (`r site_summary[site=="Osso-Luga", coord]`), is located on the Matebian foothills based on Bobonaro Scaly Clay, with an annual rainfall of `r site_summary[site=="Osso-Luga", prec]` mm and an altitude ranging between `r site_summary[site=="Osso-Luga", elev]` masl (Figure 1). The soils of the Baucau municipality are classified using the WRB soil classification [@Mantel2023] as Lithosols and Calcic Luvisols [@FAO2012]. The average soil properties differ between the two sites (Table S\@ref(tab:site-soil)); Gariuai has a clay loam texture whereas Osso-Luga has a clay texture. The two villages also have different histories. Gariuai has been inhabited for at least 50 years, while Osso-Luga was abandoned during the Indonesian occupation and was only repopulated after Timor-Leste gained independence in 2002. The AFS in Osso-Luga lay fallow during the period of abandonment. The land use of these two villages is typical of the Baucau municipality. The five AFS described in this article [see detailed description in @Cogne2024] are found in the valleys and high plateaus, while non-AFS such as rainfed rice and horticultural plots are more likely to be found in the lowlands. Other non-agricultural land uses are also present, such as limestone quarrying. Some natural forest parks are also preserved around mountain peaks, such as Matebian and Mundo Perdido. Due to the risk of flooding, villages are rarely located near rivers, and are concentrated instead near paved roads and on hilltops. 
## Typology of agroforestry systems

We used the typology of AFS described in @Cogne2024, which was developed in the same study area. This typology was based on the one proposed by Nair et al. (1993, 2021) and adapted to the local context with semi-structured interviews previously conducted in another study at the same sites [@Cogne2024]. The five AFS identified were as follows: Cropping systems with Fallow (CF), Silvopastures (SP), Young Agroforests (YA), Home Gardens (HG), and Forest Gardens (FG). The AFS are presented in more detail in the Supplemental Information (Table S\@ref(tab:afs-typology)) and in @Cogne2024.

Briefly, CF are plantations of a few non-perennial crops (e.g., maize [*Zea mays*], sweet potatoes [*Ipomoea batatas*], groundnuts [*Arachis hypogaea*], and squash gourd [*Cucurbita moschata*]) which are left fallow after one to three years of cropping. SP are vast areas (generally 200 to 500 ha) where cattle (*Bos taurus*, *Bos domesticus*, *Bubalus bubalis*) are grazed under the shade of scattered palms and trees such as coconut (*Cocos nucifera*), tamarind (*Tamarindus indica*), eucalyptus (*Eucalyptus alba*) and jujube (*Ziziphus jujuba*). YA are agricultural plots where farmers plant rows of trees, like teak (*Tectona grandis*) and mahogany (*Swietenia mahagoni*), spaced approximately 5 x 10 m apart, which are generally left to grow to become HG or FG (see Figure S\@ref(fig:seq-diagram)). HG are plots of land located close to houses where crops and livestock (poultry [*Gallus gallus domesticus*], pigs [*Sus domesticus*], cattle [*Bos taurus*, *Bos domesticus*, *Bubalus bubalis*], etc.) are kept in the shade of numerous trees and palms, like breadfruit (*Artocarpus altilis*), candlenut (*Aleurites moluccanus*) and areca palm (*Areca catechu*), all of which are put to multiple uses (wood, fruit, fibre, etc.). FG are generally former HG (over 50 years old) which have been abandoned, for example as a result of the war, and are now used mainly for gathering fruits (e.g., mango [*Mangifera indica*]) and non-timber forest products (e.g., candlenut) with limited human influence.

@@

We randomly selected three plots from each of the five traditional AFS identified in each of the two sites (30 plots in total; Figure \@ref(fig:map-sites)). Sampling plots were defined as the entire agricultural plot (area between 0.06 and 1.25 ha) for all AFS except SP plots, which were generally much larger than 1 ha. In SP plots, a 1 ha sampling plot was delineated at the centre of the agricultural plot. Tree and soil measurements (described below) were carried out between June and August 2021.

Within each sampling plot, all trees with a circumference at breast height (CBH) greater than or equal to 100 cm were identified taxonomically and their CBH and height were measured. Five subplots of 10x10 m (0.01 ha) were established at each corner and in the centre of the sampling plot (Figure S\@ref(fig:design)). Within these subplots, all trees with a CBH between 30 and 100 cm were taxonomically identified and their CBH and height were measured. Height measurements were taken with a dendrometer.

Taxonomic identification was carried out using Tetum (a national official language) vernacular names, which were then linked to scientific names using the iNaturalist and Pl\@ntNet applications, Geoffrey Hull’s monograph [@Hull2006], the “Useful tropical plants” database [@Fern2014], and the Australian Northern Territory Government’s Native Plants Herbarium [@NT2013]. The correspondence between Tetum names and scientific names is provided in the Supplemental Information (Table S\@ref(tab:tetum-names)).

In each of the 30 sample plots, a composite of three samples was taken from three of the five 10x10 m subplots (including the plot centre and the two opposite corners along the plot's longest diagonal; Figure S\@ref(fig:design)), covering the intra-plot variability and at a minimum distance of 1 m from any tree. Soils were sampled with a soil auger (Model Bor Tanah - Dormer Standard Soil Auger, Adiguna Karya Persada, Bogor, Indonesia) at depths of 0-10 cm and 20-30 cm. The sample from the 10-20 cm soil layer was omitted due to the difficulty of sampling with the 15-cm soil auger. Samples underwent standard physical and chemical analyses at CIRAD laboratories in Montpellier, France (Table S\@ref(tab:site-soil)). The soil organic carbon content of the samples collected with the soil auger was calculated as the difference between the total carbon and the inorganic carbon content of the soil. Total carbon content was determined using the Dumas dry combustion method, following the ISO 10694:1995 standard method, and measured with a thermal conductive detector (EA 1112 Elemental Analyser, Thermo Fischer Scientific, Netherlands). Soil inorganic content was determined using a calcimeter, and following the ISO 10693 standard method. Particle size distribution (proportion of clay, silt and sand) was determined using the pipette method and three classes: clay (\< 2 μm), silt (2–50 μm) and sand (50 μm - 2 mm). Soil pH was measured with a soil:deionized water ratio of 1:2.5.

In addition, at the centre of the plot, three 250 cm^3 soil samples were taken with a soil sampling cylinder (5 cm in height, 7.98 cm inner diameter, and 8.4 cm outer diameter; Eijkelkamp Soil & Water, Giesbeek, Netherlands) at depths of 2-7 cm, 12-17 cm and 22-27 cm. These depths were chosen to take into account the height of the cylinder (i.e., 5 cm), and to be associated with depths 0-10 cm, 10-20 cm and 20-30 cm, the depths at which the organic carbon content was calculated or estimated. Each sample was dried at 105°C in a classic oven and weighed repeatedly until a stable dry mass was reached, and then sieved to 2 mm. The fine fraction (\< 2 mm) was weighed to be used in the SOC stocks calculation (equation \@ref(eq:soc)).

@@

## height in cm, with a conversion factor between 0.8 and 1
data_sim_agc[GENUS %in% palm_genera, 
             HEIGHT := ifelse(
               iter == "mean", 
               HEIGHT,  ## height in m
               rnorm(length(HEIGHT), HEIGHT, HEIGHT*cvH)
             )]

## add WD data
data_wd_palms = unique(data_wd[data_wd$genus %in% palm_genera, c("genus", "meanWD", "sdWD")])
data_sim_agc = merge(
  data_sim_agc,
  data_wd_palms,
  all.x = TRUE,
  by.x = "GENUS",
  by.y = "genus"
)
data_sim_agc[GENUS %in% palm_genera, 
             WD := ifelse(iter == "mean", 
                          meanWD, 
                          rnorm(length(meanWD), meanWD,sdWD))]
# palm biomass with Goodman et al 2013 equation that includes diameter, height
# and dry mass fraction; we used an average dmf of 0.37 which 
data_sim_agc[
  GENUS %in% palm_genera, 
  AGC := ((0.463 * DIAM_130_CM^2 * HEIGHT)^0.25 * 0.55512) ^ 4	* 0.4713 *1e-3]
```

We estimated tree-level AGC stocks from the tree CBH and height measurements using the pantropical allometric equation from @Chave2014 and the R package BIOMASS [@Rejou-Mechain2017]. Each tree was assigned a mean and standard deviation of wood density at the species or genus level when at least one value was available from the Global Wood Density Database [@Zanne2009]. Other taxa were assigned a plot-level average wood density. Palm biomass was estimated using the family level allometric equation of @Goodman2013 with DBH, stem height and dry mass fraction as predictors. As we didn't have species-specific values for dry mass fraction, we used the average value of 0.463 from @Goodman2013. Biomass was converted into carbon with a factor of 0.4713 [@Thomas2012].

Belowground carbon (BGC) stocks were estimated at tree level using the root-to-shoot ratio equation from @Ledo2018, which is based on tree DBH and climatic water deficit from @Chave2014. This ratio was then multiplied by the estimated AGC stock to obtain tree-by-tree BGC stock. Plot-level AGC and BGC stocks were then obtained by summing tree-level stocks and dividing by the plot area.

```{r bgc-estimation}
# root-to-shoot ratio allometry from Ledo et al., 2018
# use CWD values from Chave et al., 2014
# download from figshare https://figshare.com/articles/dataset/Root_Shoot_global_data_individual_trees/5144164?file=8754550
# rs <- readxl::read_excel("C:/Users/piponiot-laroche/Downloads/S1Database.xlsx",
#                          sheet = "DATA")
# rs$BGB_kg <- as.numeric(rs$BGB_kg)
# rs$wd <- as.numeric(rs$wd)
# rs$dbh2 <- rs$DBH_cm^2
# lm(log(BGB_kg/AGB_kg) ~ DBH_cm + dbh2 + wd,
#    data = subset(rs, !is.na(BGB_kg) & BGB_kg != 0))

@@

@@

\begin{equation}
SOC_p = \sum_i Corg_{i,p} \cdot BDfine_{i,p}\cdot (1- \frac{Vcoarse_{i,p}}{Vcyl})\cdot (Surf \cdot \Delta d) \cdot 10^{-6}
\end{equation}

Where $ Corg_{i,p}$ is the SOC content at depth $i$ (either 0-10, 10-20 or 20-30 cm) as measured in the laboratory from the soil sampled by the auger (depths 0-10 cm and 20-30 cm) and estimated at depth 10-20 cm (see details below). $BDfine_{i,p}$ is the bulk density of the fine fraction of the soil sampled with the cylinder at depth $i$ (in g cm$^{-3}$):

$$BDfine_{i,p} = \frac{mfine_{i,p}}{Vcyl – Vcoarse_{i,p}}$$
$mfine_{i,p}$ is the mass of the fine fraction (\< 2 mm) of the sampled soil; Vcyl is the volume of the sampling cylinder (250 cm$^3$); and $Vcoarse_{i,p}$ is the volume of the coarse fraction in the sampling cylinder. $Surf = 10^8$ is the surface conversion factor from cm$^2$ to ha; $\Delta d$ is the difference in sample depths (= 10 cm); 10^{-6}$ is the conversion factor from g to Mg.

This equation could be reformulated [@Poeplau2017] as:
